Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64

### 2018-06-04 PAPER.tex
Mon Jun  4 10:09:27 EDT 2018

We illustrate the way in which the quadratic estimator, Equations \eqref{eq:qhat} and \eqref{eq:phat}, is intended to estimate the power spectrum of EoR in the presence of contamination by other terms by considering two simpler, but similar, cases of estimating the variance of data.  We have specifically chosen a toy model where the data covariance is diagonal, as indeed we expect the EoR signal to be.  We assume we have $N$ data points $\Delta_i$ which are the sum of a desired signal $\sigma_i$ and an undesired contaminant $\upsilon_i$
\begin{equation}
\Delta_i = \sigma_i + \upsilon_i
\end{equation}
with 
\begin{equation}
\langle \sigma_i \rangle = 0; \; \langle \sigma^2_i \rangle = s; \; {\rm and}~\langle \bm{\sigma \sigma}^T \rangle = s \mathbf{I}_{N \times N} \equiv \mathbf{S}
\end{equation}
and we wish to estimate $s$.  The contaminant in this case has a similar structure for its covariance, and is assumed uncorrelated with the signal
\begin{equation}
\langle \upsilon_i \rangle = 0; \; 
\langle \upsilon^2_i \rangle = u; \; 
\langle \bm{\upsilon \upsilon}^T \rangle = u \mathbf{I}_{N \times N} \equiv \mathbf{U}; {\rm and}~ 
\langle \sigma_i \upsilon_j \rangle = 0
\end{equation}
With the covariance matrix given by $\C = \mathbf{S} + \mathbf{U}$, the estimator for $s$ using the optimal quadratic estimator prescription is
\begin{equation}
\label{eq:ICWFormula}
\hat{s} = \half {\F}^{-1} \left( \bm{\Delta}^T \invC \PDeriv{\C}{s}  \invC \bm{\Delta} \right) 
\end{equation}
where $\F$ is the Fisher matrix.  
The expectation of the estimator is then
\begin{equation}
\langle \hat{s} \rangle = \frac{\langle \bm{\Delta^T \Delta} \rangle}{N} = s + u
\end{equation}
In this case, the quadratic estimator returns an entirely reasonable result, but, since the covariance of the undesired signal has the same form as the desired signal, that is,
$\PDeriv{\mathbf{S}}{s} = \PDeriv{\mathbf{U}}{u} =  \PDeriv{\C}{s}$, we find that when {\it the covariance structure of the contaminant is identical to the signal}, the information available to the quadratic estimator to distinguish the two is degenerate, and  is only possible to achieve an unbiased estimate of $s$ by using some additional information to estimate the additive bias due to $u$ and subtract it.  In fact, if one uses the standard formula \citep[e.g.][, Equation 6]{dillon14} for the additive bias
\begin{equation}
b = \half \F^{-1} \Tr\left[ (\C - \mathbf{S}) \invC \PDeriv{\C}{s} \invC \right]
\end{equation}
one indeed finds $b = g$.

Now consider a case, chosen to be very similar to the toy model in \ref{sec:toymodel},  in which the data again have an additive contaminant, now given by
\begin{equation}
\Delta_i = \sigma_i + \upsilon m_i
\end{equation}
where the properties of $\sigma_i$ are as before, but now $\upsilon$ is a random variable and $m_i$ is a fixed function of $i$ with
\begin{equation}
\langle \upsilon \rangle = 0; \;
\langle \upsilon^2 \rangle = u; \; 
 \langle \bm{\upsilon \upsilon}^T \rangle = u \mathbf{m m}^T \equiv \mathbf{U}; \;
 \mathbf{m}^T \mathbf{m} = N; \; {\rm and}~ 
\langle \sigma_i \upsilon \rangle = 0
\end{equation}
Here $\mathbf{m}$ represents a mode which is correlated across many data points, with amplitude given by  $\upsilon$.

We can again calculate the estimator given by \eqref{eq:ICWFormula} explicitly by using the Sherman-Morrison identity to invert the covariance matrix.  Defining
\begin{equation}
\xi \equiv \frac{u/s}{1+u/s}
\end{equation} 
we have
\begin{equation}
\invC  =   \frac{1}{s} \left(\I - \xi \mathbf{m m}^T \right)
\end{equation}
and the virtue of weighting by $\invC$ becomes clearer: in the limit $u \gg s$, $\xi \to 1$, and this becomes
\begin{equation}
\invC  = \frac{1}{s} \left(\I - \mathbf{m m}^T \right)
\end{equation}
where $\I - \mathbf{m m}^T$ is the projection operator, projecting out $\mathbf{m}$ from any any vector it acts on.   Proceeding to calculate the full estimator
\begin{equation}
\hat{s} = \half {\F}^{-1} \left(\bm \Delta^T \invC \PDeriv{\C}{s}  \invC \bm \Delta \right) = \frac{ \bm \Delta^T  (\I +  (\xi^2 - 2 \xi)  \mathbf{m m}^T)  \bm \Delta}{N + \xi^2 - 2 \xi} 
\end{equation}
The expectation is
\begin{equation}
\langle \hat{s} \rangle = \frac{ N(s + u + (\xi^2 - 2 \xi) 
[u + \langle(\bm{\sigma} \cdot \mathbf{m})^2 \rangle])} 
{N + \xi^2 - 2 \xi} 
\end{equation}
This again has a simple interpretation in the limit $\xi \to 1$, where
\begin{equation}
\langle \hat{s} \rangle = \frac{N}{N -1}  (s - \langle(\bm{\sigma} \cdot \mathbf{m})^2\rangle )
\end{equation}
Thus, in the limit the contaminant is large compared to $s$, and to the extent that $\bm \sigma \cdot \mathbf{m} \approx 0$, this yields a nearly unbiased estimator for $s$ and the inverse covariance weighting has completely removed the effect of the contaminant. This is indeed the desired behavior of the inverse covariance weighting, where removal of contamination with a known covariance can be suppressed by projection of the offending modes.  But even in this case, it is worth pointing out that estimator is still not unbiased in general, having both the effect of leaving in some contaminated mode for finite $\xi$ and also in the limit of some correlation between the signal and contaminant, or of mis-estimation of the contaminating mode $\mathbf{m}$.

>>>>

We illustrate the way in which the quadratic estimator, Equations \eqref{eq:qhat} and \eqref{eq:phat}, is intended to estimate the power spectrum of EoR in the presence of contamination by other terms by considering two simpler, but similar, cases of estimating the variance of data.  We have specifically chosen a toy model where the data covariance is diagonal, as indeed we expect the EoR signal to be.  We assume we have $N$ data points $\Delta_i$ which are the sum of a desired signal $\sigma_i$ and an undesired contaminant $\upsilon_i$
\begin{equation}
\Delta_i = \sigma_i + \upsilon_i
\end{equation}
with 
\begin{equation}
\langle \sigma_i \rangle = 0; \; \langle \sigma^2_i \rangle = s; \; {\rm and}~\langle \bm{\sigma \sigma}^T \rangle = s \mathbf{I}_{N \times N} \equiv \mathbf{S}
\end{equation}
and we wish to estimate $s$.  The contaminant in this case has a similar structure for its covariance, and is assumed uncorrelated with the signal
\begin{equation}
\langle \upsilon_i \rangle = 0; \; 
\langle \upsilon^2_i \rangle = u; \; 
\langle \bm{\upsilon \upsilon}^T \rangle = u \mathbf{I}_{N \times N} \equiv \mathbf{U}; {\rm and}~ 
\langle \sigma_i \upsilon_j \rangle = 0
\end{equation}
With the covariance matrix given by $\C = \mathbf{S} + \mathbf{U}$, the estimator for $s$ using the optimal quadratic estimator prescription is
\begin{equation}
\label{eq:ICWFormula}
\hat{s} = \half {\F}^{-1} \left( \bm{\Delta}^T \invC \PDeriv{\C}{s}  \invC \bm{\Delta} \right) 
\end{equation}
where $\F$ is the Fisher matrix.  
The expectation of the estimator is then
\begin{equation}
\langle \hat{s} \rangle = \frac{\langle \bm{\Delta^T \Delta} \rangle}{N} = s + u
\end{equation}
In this case, the quadratic estimator returns an entirely reasonable result, but, since the covariance of the undesired signal has the same form as the desired signal, that is,
$\PDeriv{\mathbf{S}}{s} = \PDeriv{\mathbf{U}}{u} =  \PDeriv{\C}{s}$, we find that when {\it the covariance structure of the contaminant is identical to the signal}, the information available to the quadratic estimator to distinguish the two is degenerate, and  is only possible to achieve an unbiased estimate of $s$ by using some additional information to estimate the additive bias due to $u$ and subtract it.  In fact, if one uses the standard formula \citep[e.g.][, Equation 6]{dillon14} for the additive bias
\begin{equation}
b = \half \F^{-1} \Tr\left[ (\C - \mathbf{S}) \invC \PDeriv{\C}{s} \invC \right]
\end{equation}
one indeed finds $b = u$.

Now consider a case, chosen to be very similar to the toy model in \ref{sec:toymodel},  in which the data again have an additive contaminant, now given by
\begin{equation}
\Delta_i = \sigma_i + \upsilon m_i
\end{equation}
where the properties of $\sigma_i$ are as before, but now $\upsilon$ is a random variable and $m_i$ is a fixed function of $i$ with
\begin{equation}
\langle \upsilon \rangle = 0; \;
\langle \upsilon^2 \rangle = u; \; 
 \langle \bm{\upsilon \upsilon}^T \rangle = u \mathbf{m m}^T \equiv \mathbf{U}; \;
 \mathbf{m}^T \mathbf{m} = 1; \; {\rm and}~ 
\langle \sigma_i \upsilon \rangle = 0
\end{equation}
Here $\mathbf{m}$ represents a mode which is correlated across many data points (i.e., we are assuming $\mathbf{U}$ need {\it not} be diagonal), with amplitude given by  $\upsilon$.  The normalization of $\mathbf{m}$ is a matter of convention, and can be absorbed in the variance $u$; the choice above will be convenient for understanding the limiting case $u \gg s$.

We can again calculate the estimator given by \eqref{eq:ICWFormula} explicitly by using the Sherman-Morrison identity to invert the covariance matrix.  Defining
\begin{equation}
\xi \equiv \frac{u/s}{1+u/s}
\end{equation} 
we have
\begin{equation}
\invC  =   \frac{1}{s} \left(\I - \xi \mathbf{m m}^T \right)
\end{equation}
Proceeding to calculate the full estimator
\begin{equation}
\hat{s} = \half {\F}^{-1} \left(\bm \Delta^T \invC \PDeriv{\C}{s}  \invC \bm \Delta \right) = \frac{ \bm \Delta^T  (\I +  (\xi^2 - 2 \xi)  \mathbf{m m}^T)  \bm \Delta}{N + \xi^2 - 2 \xi} 
\end{equation}
The expectation is
\begin{equation}
\langle \hat{s} \rangle = 
s + \frac{1 - 2 \xi + \xi^2}{N + - 2 \xi + \xi^2} u 
\end{equation}
It is worth observing immediately that there is no multiplicative bias on $s$, and that the additive bias is strictly $< u/N$.

An instructive limit is $u \gg s$, $\xi \to 1$, in which case the the virtue of weighting by $\invC$ becomes clearer: it becomes
\begin{equation}
\invC  = \frac{1}{s} \left(\I - \mathbf{m m}^T \right)
\end{equation}
where $\I - \mathbf{m m}^T$ is the projection operator, projecting out $\mathbf{m}$ from any any vector it acts on, and further, the linear bias tends to 0 as $\xi \to 1$, i.e., the projection is ``perfect'' and not ``undone'' by the Fisher matrix normalization.

This is indeed the desired behavior of the inverse covariance weighting, where removal of contamination with a known covariance can be suppressed by a kind of projection of the offending modes.  But even in this case, it is worth pointing out that estimator still has a linear bias for finite $u$.  We also assume that the contaminating mode $\mathbf{m}$ is known perfectly; the next appendix takes up the case where the modes are estimated from the data.
